---
title: "STATS5099 Data Mining and Machine Learning"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
biblio-style: apalike
link-citations: yes
---

```{r setup, include = FALSE}
library(webexercises)
library(tinytex)
library(dplyr)
library(knitr)
library(kableExtra)
knitr::opts_chunk$set(echo=TRUE, eval=TRUE, warning=FALSE, message=FALSE)

library(dendextend)
```


```{r include=FALSE, echo=FALSE}
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```


# Welcome to DMML Lab 8

In week 8, we learned about performing clustering by using hierarchical agglomerative clustering (HAC), visualising HAC via dendrograms, and selecting the optimal number of clusters according to the silhouette width. 

The command for hierarchical agglomerative clustering is `hclust`. Setting the method argument selects the linkage method. It takes as its first argument a distance matrix. The distance matrix can be created using the command `dist` (default method is Euclidean distance).
```{r eval=FALSE}
d <- dist(data,method="euclidian")
hiclust <- hclust(d, method="single")
```

To plot the resulting dendrogram, the plot command is used on the `hclust` fitted object. If the dendrogram suggests $g$  clusters, we could determine the cluster assignment by using the command `cutree` within the `dendextend` package with argument `k` set to `g`. Alternatively, if the dendrogram should be cut at height $l$, we could set the argument `h` to `l`. The result will be a vector with length equal to the number of observations, each entry identifying which cluster the corresponding object is assigned to.
```{r eval=FALSE}
plot(hiclust)
library(dendextend)
clusters <- cutree(hiclust, k=3)
clusters <- cutree(hiclust, h=20)
```

It is sometimes also helpful to visualise the data giving different colours to observations belonging to different clusters. This can be done by using the function `color_branches` within the `dendextend` package. 
```{r eval=FALSE}
library(dendextend)
dend <- as.dendrogram(hiclust,hang=0)
dend_cut_3clusters <- color_branches(dend,k=3) #3 clusters
plot(dend_cut_3clusters, horiz=TRUE)
```

To determine the optimal number of clusters and/or assess the performance of clustering algorithms, silhouette plots can be used. The silhouette value ranges from −1 to 1, where a high value indicates that the object is well matched to its own cluster and poorly matched to neighbouring clusters. In `R`, we could either use the function `silhouette` after clustering. 
```{r eval=FALSE}
d <- dist(data,method="euclidian")
hiclust <- hclust(d, method="single")
clusters <- cutree(hiclust, k=2)
plot(silhouette(clusters), dist(data), col=1:2)
```

The function `fviz_nbclust` from the package `factoextra` can produce a plot of the average silhouette width against the number of clusters, which helps determine the optimal number of clusters. We would have to use the `FUN=hcut` argument to specify hierarchical clustering.
```{r eval=FALSE}
library(factoextra)
ggplot_fviz <-fviz_nbclust(USArrests,FUN=hcut,method = "silhouette")
ggplot_fviz
```

